---
title: "Introduction to single-cell RNA-seq analysis"
subtitle: "Cluster marker genes"
output:
  html_document:
    toc: yes
    toc_float: true
    number_sections: true
    code_folding: show 
    css: ../css/boxes.css
---


```{r setup, echo=FALSE, message=FALSE, warning=FALSE, purl=FALSE}
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE, cache=TRUE)
set.seed(123)
```


# Introduction

In order to aid the interpretation of the clustering results that we covered in
the [previous section](08_Clustering.html), it is helpful to identify genes that
contribute to the separation of cells into those clusters.

The main approach to achieve this, is to identify genes that are differently
expressed between clusters. These may be, for example, exclusively expressed in
a single cluster or perhaps differentiate between a few different clusters.
There are different methods to identify expression differences between clusters:
using mean expression level, or the ranking of the gene by its expression, or
the proportions of cells that express the gene.

Our main objective in this section is to cover some of the methods that can be used to achieve this goal, and obtain a summary table of results.
As always, the [OSCA chapter on marker detection](https://bioconductor.org/books/release/OSCA.basic/marker-detection.html)
contains additional detail and advice.

# Load data

```{r load_packages}
library(scater)
library(scran)
library(tidyverse)
library(patchwork)
```

We will use the data set generated in the clustering session. This contains 7
samples from the Caron data set. For the purposes of these materials, in the
interests of time, each sample has been downsampled to only contain 500 cells.

```{r load_data}
sce <- readRDS("R_objects/Caron_clustered.500.rds")
```

Note that we have also replaced the gene names (rownames) of our objects to use
common gene names instead of Ensembl IDs. This was done using the
`uniquifyFeatureNames()` function, which makes this safely by avoiding duplicate gene
names.

```{r look_at_rownames, purl=FALSE}
rownames(sce)[11:20]
```


For this demonstration we will investigate the clustering generated using the
Leiden clustering algorithm with *k* set to 25. The results of this clustering
have been added to a column in the `colData` called **"label"** using the 
`colLabels()` accessor. The advantage of this is that should we later decide to
use a different clustering or labels (such as cell types), we can simply change 
the contents of this column and there would be no need to modify any subsequent 
code.

Double check that the "label" column contains the clustering that we are
interested in:

```{r check_label_column}
all(sce$k.25_cluster.fun.leiden == sce$label)
```

To remind ourselves, we can visualise the clusters on a UMAP:

```{r clusters_umap}
plotReducedDim(sce, 
               dimred = "UMAP_corrected",
               colour_by = "label", 
               text_by = "label")
```

Our objective is to identify genes that distinguish these clusters from one
another - "cluster marker genes". Intuitively we hope that the clusters relate
to specific cell populations, and therefore we are trying to find genes that
will allow us to identify the cell types for each cluster.


# Identifying cluster marker genes

Although we have defined our clusters based on the batch-corrected expression
values, these should **not** be used for for gene-based analyses like marker
gene detection. Instead, we should **use the uncorrected (normalised) expression
values for differential expression between clusters**. This is because data
integration algorithms bring cells together based on their overall gene
expression, but for each gene individually the data transformation may introduce
artificial agreement between batches, which is not ideal for gene-level
differential analysis. Furthermore, the severity of these biases is dependent on
the parameters used for data integration (such as the number of nearest
neighbours in the `fastMNN()` method).

Valid assays to use in gene based differential analysis tests are the normalised
counts obtained from the deconvolution method (using
`scuttle::computePooledFactors()` + `scuttle::logNormCounts()`) or from the
variance stabilising transformation method (using `sctransform::vst()`). In our
SCE object, we have the normalised counts in the "logcounts" assay, which we can
access with `assay(sce, "logcounts")` (or using the shortcut `logcounts(sce)`).


## Pairwise Cluster Comparisons

The basic approach for marker gene identification across clusters is to perform
statistical tests for each gene between every pair of clusters. The
`scoreMarkers()` function can do this for us, while accounting for known factors
(aka "blocking factors" or "blocks"), such as sample batch. 

The `scoreMarkers()` function outputs a _list_ of `DataFrame` objects, one for each
cluster compared to all others. However, note that the blocking assumes that
each pair of clusters is present in at least one of the blocks. If there are two
clusters which are not both present in at least one block (in our case
Samples), then that pairwise comparison will by necessity be omitted.

By default the `scoreMarkers()` function will use the log-normalised counts as
stored in the "logcounts" assay slot of the single cell object, so there is no 
need for us to specify it.

```{r score_markers}
markers <- scoreMarkers(sce, 
                        groups = sce$label, 
                        block = sce$SampleName)
```

The returned object is a list of the same length as the number of clusters. 
We can access the results for a particular cluster thus:

```{r score_markers_results_cluster_11}
c11_markers <- as.data.frame(markers[["11"]])
head(c11_markers)
```

This DataFrame contains the results for cluster 10. 
The first four columns contain summary statistics:

* **self.average** - the mean log-expression in the cluster of interest
* **other.average** - the grand mean across all other clusters
* **self.detected** - the proportion of cells with detected expression in the
cluster of interest
* **other.detected** - the mean detected proportion across all other clusters.

The remaining columns contain summaries of three scores from the pairwise
comparisons. The three scores are:

* **logFC.cohen** - "Cohen's *d*" - this is the log fold change of mean gene
expression standardized by the average standard deviation across the groups. 
This can be interpreted in a similar way to log fold change in that a positive 
value indicates upregulation in the cluster of interest.
* **AUC** - "Area Under the Curve" - this quantifies the ability to distinguish
between two gene expression distributions. It can be interpreted as the
likelihood that any random cell in the cluster of interest will have a higher
expression of the gene than any random cell in the other cluster. It ranges from
0 to 1, where 1 can be interpreted as upregulation, 0 downregulation, and 0.5 as
no difference.
* **logFC.detected** - this is the log fold change in the proportion of cells in
which the gene is detected in the cluster of interest, versus the proportion of
cells in which the gene is detected in the other cluster. Positive values
indicate that the gene is detected in more cells in the cluster of interest than
the other cluster. Note, this takes no account of the magnitude of the gene
expression, instead this metric helps to identify presence/absence differences
in gene expression between clusters.

More detail on the differences between these effect size scores can be found in
the ["Advanced" Marker detection chapter of the OSCA
book](http://bioconductor.org/books/3.15/OSCA.advanced/marker-detection-redux.html#properties-of-each-effect-size).

Whilst all the pairwise scores can be retrieved by adding the argument
`full.stats=TRUE` to `scoreMarkers()`, by default this function returns 5
summary statistics for each score:

* **mean.X** - this is the mean of the score across all pairwise comparisons. It
gives the relative expression of the gene versus the average of the other
clusters.
* **min.X** - this is the minimum score obtained across all pairwise
comparisons. This is the most stringent summary statistic for detecting
upregulated genes, if the score is high, then the gene is upregulated in the
cluster of interest relative to *all* other clusters. Conversely, if the minimum
is low, then the gene is downregulated relative to *at least one* other cluster.
* **median.X** - this is the median of the score across all pairwise
comparisons. It is more robust to outliers than the mean. Whilst this is
beneficial in avoiding the situation where the effect size is very large in only
a small number of comparisons, it may conversely be detrimental to marker gene
selection if there are small number of clusters with very similar effect sizes,
as these will be effectively ignored.
* **max.X** - this is the maximum score obtained across all pairwise
comparisons. This is the least stringent summary statistic for detecting
upregulated genes as a high score only indicates that the gene is upregulated in
the cluster of interest relative to *at least one* other clusters. Conversely,
if the maximum is low, then the gene is downregulated relative to *all* other
clusters.
* **rank.X** - This is the minimum ranking ("min-rank") of that gene by that
score across all clusters. For each pairwise comparison the genes are ranked
according to the score, this summary provides the lowest rank for that gene
across all comparisons. Essentially, a gene with a low "min-rank" will be one of
the best genes according to that score in at least one comparison.

The choice of the summary used for ranking will effect the stringency of the
selection. See the [the OSCA books "Basic" chapter on Marker gene
detection](http://bioconductor.org/books/3.15/OSCA.basic/marker-detection.html#summarizing-pairwise-effects)
for further discussion of these different summaries. In general the mean and
median make reasonable defaults for most applications. In practice, the minimum
and maximum are most helpful for diagnosing discrepancies between the mean and
median, rather than being used directly for ranking.

Selecting genes based on a given min-rank, say *5*, is useful as it will
generate a list of genes that is the union of genes with a rank of 5 or less for
any pairwise comparison. This will ensure we get at least 5 genes that
distinguish the cluster of interest from all other clusters.

For example using the min-rank for Cohen's *d* on cluster 11 yields 20 marker
genes: 

```{r top_genes_cluster_11}
c11_markers %>% 
  select(contains("cohen")) %>%
  filter(rank.logFC.cohen <= 5) %>%
  arrange(rank.logFC.cohen)
```

As we suspected, cluster 11 is likely to contain monocytes, based on the
expression of _CST3_ and _TYROBP_, for example.

We can visualise the expression of some of the other top-ranked genes: 

```{r plot_UMAP_LYZ, fig.width=12}
p1 <- plotReducedDim(sce, 
                     dimred = "UMAP_corrected",
                     colour_by = "LYZ", 
                     text_by = "label")

p2 <- plotExpression(sce, features = "LYZ", x = "label")

p1 + p2
```

Based on the expression of _CD3D_, which is a known marker for T cells, it seems
likely that cells in clusters 6 and 7 are T cells.

```{r exercise_1_CD3D_expression}
plotReducedDim(sce, 
               dimred = "UMAP_corrected",
               colour_by = "CD3D", 
               text_by = "label")

plotExpression(sce, 
               features = "CD3D", 
               x = "label")
```

We would like to confirm this by identifying other genes that differentiate
these two clusters from the rest of the cells.

Start by extracting the marker results for cluster 6 from our list:

```{r exercise_1_answer_a, purl=FALSE}
c6_markers <- as.data.frame(markers[["6"]])
```

In our case, we chose to then filter our table with two criteria to find genes
that:

* are in the top 2 ranking for being mostly detected in this cluster (i.e. a presence/absence criteria) 

AND

* have a Cohen's _d_ in the top 10 (i.e. the average expression difference
should be high, relative to the gene's variance)

```{r exercise_1_answer_b, purl=FALSE}
c6_top_genes <- c6_markers %>% 
  filter(rank.logFC.detected <= 2 & rank.logFC.cohen <= 10) %>% 
  rownames()
c6_top_genes
```

We obtain `r length(c6_top_genes)` genes with these criteria.
Finally, we visualize the expression of these genes:

```{r exercise_1_answer_c, purl=FALSE}
plotExpression(sce, 
               features = c6_top_genes, 
               x = "label")
```

We can see that all of these genes behave in a similar way to our known
marker _CD3D_, suggesting they are good markers for these cell types. 
This makes sense as _CD3D_ and _CD3E_ encode T cell surface proteins.
[_IL32_](https://en.wikipedia.org/wiki/Interleukin_32) encodes a cytokine, which
is often associated with cancer and _BCL11B_ is a transcription factor that has
been linked with T-Cell malignancy.
This result could open an avenue for further investigation in this study.

## Heatmap of marker genes

We have already seen how we can use the `plotExpression()` function to visualise
the distribution of expression in our data between clusters. We have also seen
how to use `plotReducedDim()` to visualise a gene's expression on the projected
reduced dimensionality space.

Another useful type of visualisation is to use **heatmaps** to show the
expression of these genes of interest. 
We will demonstrate this using the top marker genes for cluster 11.

```{r cluster_11_top_genes}
c11_top_genes <- c11_markers %>% 
  filter(rank.logFC.cohen <= 5) %>% 
  rownames()
```

We can make generate a heatmap showing the expression in individual cells using
the function `plotHeatmap()`. We will arrange the columns (cells) so that cells
from the same cluster and cells from the same samplegroup are grouped together.

```{r cluster_11_heatmap_by_cells}
plotHeatmap(sce, 
            features = c11_top_genes,
            order_columns_by = c("label", "SampleGroup"))
```

Alternatively, we can summarise the expression across sample goups and generate
a heatmap showing the average expression across cells within each group using
the function `plotGroupedHeatmap()`. We can specify any factors causing batch
effects using the `block` arguments and the batch effects will be regressed out
of the averages.

```{r cluster_11_heatmap_by_block}
plotGroupedHeatmap(sce, 
                   features = c11_top_genes,
                   group = "label",
                   block = "SampleGroup")
```

In both cases, the colour scale of expression is showing the logcounts in their
original scale. However, for this kind of visualisation, it may sometimes be
useful to _scale_ the data (aka Z-score), which brings all the genes to the same
relative scale.

```{r heatmap_with_z_score}
plotHeatmap(sce, 
            features = c11_top_genes,
            order_columns_by = c("label", "SampleGroup"),
            scale = TRUE, 
            center = TRUE,
            zlim = c(-3, 3))

plotGroupedHeatmap(sce, 
                   features = c11_top_genes,
                   group = "label",
                   block = "SampleGroup",
                   scale = TRUE, 
                   center = TRUE,
                   zlim = c(-3, 3))
```

In this case, the colour scale can be interpreted as the number of standard
deviations above/below the mean of that gene across all cells.

Another useful visualisation is to use **dot plots** of expression that show both the average gene expression (as a colour scale) and the number of cells in which the gene is detected (as the size of the points).  
We can generate such a plot using the `plotDots()` function: 

```{r dot_plot}
plotDots(sce, 
         features = c11_top_genes,
         group = "label", 
         block = "SampleGroup",
         scale = TRUE, center = TRUE, zlim = c(-3, 3))
```


## Adjusting the log-fold change

The AUC and Cohen's *d* scores incorporate both the gene expression differences
between the clusters and the variance in gene expression scores within each
cluster. If a gene has low variance, it is possible that it will be ranked
highly even if the magnitude of the difference between the clusters is low.
These genes will not necessarily make good marker genes. It may therefore be
desirable to favour the detection of genes with larger log-fold changes. 

For example, in the results from cluster 11, the gene _SNX10_ had a min-rank for Cohen's *d* of 5:

```{r c11_top_markers}
c11_markers["SNX10", ] %>% 
  select(min.logFC.cohen, max.logFC.cohen, rank.logFC.cohen)
```

However, we can also see that its LFC goes from 0.3 to 7, which is a large range. 
Looking at its expression, we can see what might be going on:

```{r plot_snx10_expression}
plotExpression(sce,
               features = "SNX10",
               x = "label")
```

This gene has very low variation in expression in some clusters (because it's lowly detected), and because Cohen's _d_ measures average differences _scaled by variance_, the gene comes up as having a high value for that metric in some comparisons.

To make our analysis more restrictive, we can instead indicate to the `scoreMarkers()` function what is the minimum LFC threshold we want to use to consider a gene for ranking. 
For example, a LFC > 2:

```{r score_markers_with_lfc_threshold}
markers_lfc <- scoreMarkers(sce,
                           groups = sce$label,
                           block = sce$SampleName,
                           lfc = 2)

c11_markers_lfc2 <- as.data.frame(markers_lfc[["11"]])
```

Now, _SNX10_'s rank dropped substantially: 

```{r c12_FLT3_threholded_rank}
c11_markers_lfc2["SNX10",  c("rank.logFC.cohen")]
```

In conclusion, using an LFC threshold will change the ranking of the genes, to favour those genes that have highest LFC, even if they have higher variance in expression. 

:::note

**Obtaining p-values from marker gene analysis**

You will notice that we did not make use of p-values in this analysis. 
The rationale for this is explained in the OSCA book:

> Given that scoreMarkers() already reports effect sizes, it is tempting to take
the next step and obtain p-values for the pairwise comparisons. Unfortunately,
the p-values from the relevant tests cannot be reliably used to reject the null
hypothesis. This is because DE analysis is performed on the same data used to
obtain the clusters, which represents “data dredging” (also known as fishing or
data snooping). The hypothesis of interest - are there differences between
clusters? - is formulated from the data, so we are more likely to get a positive
result when we re-use the data set to test that hypothesis.

The main thing to remember is that, in practice, this is a valid approach to
help us annotate groups of cells based on the expression of genes with known
cell-specificity and to find new interesting genes for further experiments and
validation (e.g. using microscopy or qPCR).
In other words, identifying cluster marker genes should be taken as a way to
generate new hypothesis from our data, rather than a valid statistical model to
test for differential expression between cell types.

:::


# Session information

<details><summary>`sessionInfo()`</summary>

```{r, purl=FALSE}
devtools::session_info()
```

</details>
